---
title: "Final Tutorial"
author: "Dani Holthaus"
date: "2024-04-30"
output: html_document
---

#### START OF COMBINED ANALYSIS ####

# Load the packages
```{r}
library(pacman)
p_load(sf,sp,raster,mgcv,plotrix,gstat,tidyverse,terra,geodata,predicts, RColorBrewer)
```


# Load the datasets
```{r}
#Assign CRS objects
wgs84 <- "+proj=longlat +datum=WGS84 +no_defs +type=crs"
albers <- "+proj=aea +lat_0=40 +lon_0=-96 +lat_1=20 +lat_2=60 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs +type=crs"

# Download bpen & bimp occurence dataset
df1 <- read.csv("https://raw.githubusercontent.com/abraham-arbelaez/SDMs-bee-project/main/BBOccurenceRecords_Verified_KS.csv")
df1.bpen <- df1 %>%
  filter(YearObserved %in% c("2020", "2021", "2022", "2023")) %>%
  filter(ScientificName == "Bombus pensylvanicus (De Geer, 1773)") %>%
  dplyr::select(ID, YearObserved, Latitude, Longitude)# Keep only the data on current bpen records (2020-2023)

df1.bimp <- df1 %>%
  filter(YearObserved %in% c("2020", "2021", "2022", "2023")) %>%
  filter(ScientificName == "Bombus impatiens Cresson, 1863") %>%
  dplyr::select(ID, YearObserved, Latitude, Longitude)

df1.bgris <- df1 %>%
  filter(YearObserved %in% c("2020", "2021", "2022", "2023")) %>%
  filter(ScientificName == "Bombus griseocollis (De Geer, 1773)") %>%
  dplyr::select(ID, YearObserved, Latitude, Longitude)

df1.bfrat <- df1 %>%
  filter(YearObserved %in% c("2020", "2021", "2022", "2023")) %>%
  filter(ScientificName == "Bombus fraternus (Smith, 1854)") %>%
  dplyr::select(ID, YearObserved, Latitude, Longitude)


# Download great plains bumble bee atlas survey dataset (filter to gpbba only & target species)
df1.bba <- read.csv("https://raw.githubusercontent.com/abraham-arbelaez/SDMs-bee-project/main/BBW_KansasRecords_5April2024_Holthaus_KState.csv")

df1.bba.total <- df1.bba %>%
  filter(datasetID %in% c("Great Plains Bumble Bee Atlas", "Bumble Bee Atlas", 
                          "Nebraska Bumble Bee Atlas", "Missouri Bumble Bee Atlas"))
df1.BPEN <- df1.bba.total %>%
  filter(vernacularName == "American bumble bee")
df1.BIMP <- df1.bba.total %>%
  filter(vernacularName == "Common eastern bumble bee")
df1.BGRIS <- df1.bba.total %>%
  filter(vernacularName == "Brown-belted bumble bee")
df1.BFRAT <- df1.bba.total %>%
  filter(vernacularName == "Southern plains bumble bee")


# Download KS boundary shapefile
sf.us <- st_read("cb_2015_us_state_20m.shp",quiet = TRUE)
sf.kansas <- sf.us[48,6] # filter to KS
sf.kansas <- as(sf.kansas, 'Spatial') #rese


# Download PRISM climate data (monthly, annual values, 4km resolution)
precip <- raster("prism/PRISM_ppt_stable_4kmM3_2022_bil/PRISM_ppt_stable_4kmM3_2022_bil.bil")
tmax <- raster("prism/PRISM_tmax_stable_4kmM3_2022_bil/PRISM_tmax_stable_4kmM3_2022_bil.bil")
tmin <- raster("prism/PRISM_tmin_stable_4kmM3_2022_bil/PRISM_tmin_stable_4kmM3_2022_bil.bil")
tmean <- raster("prism/PRISM_tmean_stable_4kmM3_2022_bil/PRISM_tmean_stable_4kmM3_2022_bil.bil")

precip <- projectRaster(precip, crs = wgs84, method = "bilinear")
tmax <- projectRaster(tmax, crs = wgs84, method = "bilinear")
tmin <- projectRaster(tmin, crs = wgs84, method = "bilinear")
tmean <- projectRaster(tmean, crs = wgs84, method = "bilinear")

# Download land cover raster for 2022 from Cropscape
rl.nlcd2022 <- raster("CDL_2022_20.tif")
plot(rl.nlcd2022)

rl.nlcd.ag <- raster("C:/Users/dyhol/OneDrive - Kansas State University/BBSDM/BBSDM_KS_2022/rl.nlcd.ag.tif")
rl.nlcd.dev <- raster("C:/Users/dyhol/OneDrive - Kansas State University/BBSDM/BBSDM_KS_2022/rl.nlcd.dev.tif")
rl.nlcd.grass <- raster("C:/Users/dyhol/OneDrive - Kansas State University/BBSDM/BBSDM_KS_2022/rl.nlcd.grass.tif")
rl.nlcd.wet <- raster("C:/Users/dyhol/OneDrive - Kansas State University/BBSDM/BBSDM_KS_2022/rl.nlcd.wet.tif")
rl.nlcd.wood <- raster("C:/Users/dyhol/OneDrive - Kansas State University/BBSDM/BBSDM_KS_2022/rl.nlcd.wood.22.tif")

#rl.nlcd2022 <- projectRaster(rl.nlcd2022, crs = wgs84, method = "ngb") #WGS84


```


# Data Wrangling - Bumble bee records
```{r}
# Make XY df
df.bpen <- data.frame(long = df1.bpen$Longitude, lat = df1.bpen$Latitude) #long, lat (XY); class = df
df.bimp <- data.frame(long = df1.bimp$Longitude, lat = df1.bimp$Latitude)
df.bgris <- data.frame(long = df1.bgris$Longitude, lat = df1.bgris$Latitude)
df.bfrat <- data.frame(long = df1.bfrat$Longitude, lat = df1.bfrat$Latitude)
df.BPEN <- data.frame(long = df1.BPEN$decimalLongitude, lat = df1.BPEN$decimalLatitude)
df.BIMP <- data.frame(long = df1.BIMP$decimalLongitude, lat = df1.BIMP$decimalLatitude)
df.BGRIS <- data.frame(long = df1.BGRIS$decimalLongitude, lat = df1.BGRIS$decimalLatitude)
df.BFRAT <- data.frame(long = df1.BFRAT$decimalLongitude, lat = df1.BFRAT$decimalLatitude)

# Make SpatialPoints data frame
pts.bpen <- data.frame(long = df1.bpen$Longitude,lat = df1.bpen$Latitude)
coordinates(pts.bpen) =~ long + lat
proj4string(pts.bpen) <- CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0")

pts.bimp <- data.frame(long = df1.bimp$Longitude, lat = df1.bimp$Latitude)
coordinates(pts.bimp) =~ long + lat
proj4string(pts.bimp) <- CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0")

pts.bgris <- data.frame(long = df1.bgris$Longitude, lat = df1.bgris$Latitude)
coordinates(pts.bgris) =~ long + lat
proj4string(pts.bgris) <- CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0")

pts.bfrat <- data.frame(long = df1.bfrat$Longitude, lat = df1.bfrat$Latitude)
coordinates(pts.bfrat) =~ long + lat
proj4string(pts.bfrat) <- CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0")

pts.BPEN <- data.frame(long = df1.BPEN$decimalLongitude, lat = df1.BPEN$decimalLatitude)
coordinates(pts.BPEN) =~ long + lat
proj4string(pts.BPEN) <- CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0")

pts.BIMP <- data.frame(long = df1.BIMP$decimalLongitude, lat = df1.BIMP$decimalLatitude)
coordinates(pts.BIMP) =~ long + lat
proj4string(pts.BIMP) <- CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0")

pts.BGRIS <- data.frame(long = df1.BGRIS$decimalLongitude, lat = df1.BGRIS$decimalLatitude)
coordinates(pts.BGRIS) =~ long + lat
proj4string(pts.BGRIS) <- CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0")

pts.BFRAT <- data.frame(long = df1.BFRAT$decimalLongitude, lat = df1.BFRAT$decimalLatitude)
coordinates(pts.BFRAT) =~ long + lat
proj4string(pts.BFRAT) <- CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0")


```

# Data Wrangling -- Create Quadratures
```{r}
set.seed(420)
quad <- spsample(sf.kansas, n = 8000, type = "random")
plot(quad)

# Make XY df
df.quad <- data.frame(long = coordinates(quad)[,1],lat = coordinates(quad)[,2]) #long, lat (XY), class = df

# Make SpatialPoints df
quad.bimp <- data.frame(long = coordinates(quad)[,1],lat = coordinates(quad)[,2])
coordinates(quad.bimp) =~ long + lat
proj4string(quad.bimp) <- CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0") #SpatialPoints

```


# Data Wrangling -- Land Cover (assign categories)-- SOMETHING WRONG HERE!!!!
```{r}

# Assign 1 to grassland/pasture land cover categories
rl.nlcd.grass <- rl.nlcd2022
rl.nlcd.grass[] <- ifelse(rl.nlcd.grass[]==176,1,0) #0 if not grassland


# Assign 1 to developed (all intensities) land cover categories
rl.nlcd.dev <- rl.nlcd2022
rl.nlcd.dev[] <- ifelse(rl.nlcd.dev[] %in% c(82, 121:124), 1,0)


# Assign 1 to water/wetland land categories
rl.nlcd.wet <- rl.nlcd2022
rl.nlcd.wet[] <- ifelse(rl.nlcd.wet[] %in% c(83, 87, 92, 111, 129, 190), 1,0)


# Assign 1 to forest/shrubland land categories
rl.nlcd.wood <- rl.nlcd2022
rl.nlcd.wood[] <- ifelse(rl.nlcd.wood[] %in% c(63:64,141:143, 152), 1,0)



# Assign 1 to agriculture land categories
rl.nlcd.ag <- rl.nlcd2022
rl.nlcd.ag[] <- ifelse(rl.nlcd.ag[] %in% c(1:60,66:77,204:254), 1,0)


### Save new raster layers (for quicker import)
#writeRaster(rl.nlcd.ag, "rl.nlcd.ag.tif")
#writeRaster(rl.nlcd.dev, "rl.nlcd.dev.tif")
#writeRaster(rl.nlcd.grass, "rl.nlcd.grass.22.tif")
#writeRaster(rl.nlcd.wet, "rl.nlcd.wet.tif")
#writeRaster(rl.nlcd.wood, "rl.nlcd.wood.22.tif")

```


# Land cover extracts
```{r}
start_time <- Sys.time()
# Grass percentage
df.bpen$grass.perc <- unlist(lapply(raster::extract(rl.nlcd.grass,pts.bpen,buffer=2000),mean))
df.bimp$grass.perc <- unlist(lapply(raster::extract(rl.nlcd.grass,pts.bimp,buffer=2000),mean))
df.bgris$grass.perc <- unlist(lapply(raster::extract(rl.nlcd.grass,pts.bgris,buffer=2000),mean))
df.bfrat$grass.perc <- unlist(lapply(raster::extract(rl.nlcd.grass,pts.bfrat,buffer=2000),mean))
df.BPEN$grass.perc <- unlist(lapply(raster::extract(rl.nlcd.grass,pts.BPEN,buffer=2000),mean))
df.BIMP$grass.perc <- unlist(lapply(raster::extract(rl.nlcd.grass,pts.BIMP,buffer=2000),mean))
df.BGRIS$grass.perc <- unlist(lapply(raster::extract(rl.nlcd.grass,pts.BGRIS,buffer=2000),mean))
df.BFRAT$grass.perc <- unlist(lapply(raster::extract(rl.nlcd.grass,pts.BFRAT,buffer=2000),mean))
df.quad$grass.perc <- unlist(lapply(raster::extract(rl.nlcd.grass,quad.bimp,buffer=2000),mean)) #2km buffer around survey location, use maximum likelihood to estimate buffer parameter if you are concerned

# Dev percentage
df.bpen$dev.perc <- unlist(lapply(raster::extract(rl.nlcd.dev,pts.bpen,buffer=2000),mean))
df.bimp$dev.perc <- unlist(lapply(raster::extract(rl.nlcd.dev,pts.bimp,buffer=2000),mean))
df.bgris$dev.perc <- unlist(lapply(raster::extract(rl.nlcd.dev,pts.bgris,buffer=2000),mean))
df.bfrat$dev.perc <- unlist(lapply(raster::extract(rl.nlcd.dev,pts.bfrat,buffer=2000),mean))
df.BPEN$dev.perc <- unlist(lapply(raster::extract(rl.nlcd.dev,pts.BPEN,buffer=2000),mean))
df.BIMP$dev.perc <- unlist(lapply(raster::extract(rl.nlcd.dev,pts.BIMP,buffer=2000),mean))
df.BGRIS$dev.perc <- unlist(lapply(raster::extract(rl.nlcd.dev,pts.BGRIS,buffer=2000),mean))
df.BFRAT$dev.perc <- unlist(lapply(raster::extract(rl.nlcd.dev,pts.BFRAT,buffer=2000),mean))
df.quad$dev.perc <- unlist(lapply(raster::extract(rl.nlcd.dev,quad.bimp,buffer=2000),mean)) #2km buffer around survey location, use maximum likelihood to estimate buffer parameter if you are concerned

# Wet percentage
df.bpen$wet.perc <- unlist(lapply(raster::extract(rl.nlcd.wet,pts.bpen,buffer=2000),mean))
df.bimp$wet.perc <- unlist(lapply(raster::extract(rl.nlcd.wet,pts.bimp,buffer=2000),mean))
df.bgris$wet.perc <- unlist(lapply(raster::extract(rl.nlcd.wet,pts.bgris,buffer=2000),mean))
df.bfrat$wet.perc <- unlist(lapply(raster::extract(rl.nlcd.wet,pts.bfrat,buffer=2000),mean))
df.BPEN$wet.perc <- unlist(lapply(raster::extract(rl.nlcd.wet,pts.BPEN,buffer=2000),mean))
df.BIMP$wet.perc <- unlist(lapply(raster::extract(rl.nlcd.wet,pts.BIMP,buffer=2000),mean))
df.BGRIS$wet.perc <- unlist(lapply(raster::extract(rl.nlcd.wet,pts.BGRIS,buffer=2000),mean))
df.BFRAT$wet.perc <- unlist(lapply(raster::extract(rl.nlcd.wet,pts.BFRAT,buffer=2000),mean))
df.quad$wet.perc <- unlist(lapply(raster::extract(rl.nlcd.wet,quad.bimp,buffer=2000),mean)) #2km buffer around survey location, use maximum likelihood to estimate buffer parameter if you are concerned

# Wood percentage
df.bpen$wood.perc <- unlist(lapply(raster::extract(rl.nlcd.wood,pts.bpen,buffer=2000),mean))
df.bimp$wood.perc <- unlist(lapply(raster::extract(rl.nlcd.wood,pts.bimp,buffer=2000),mean))
df.bgris$wood.perc <- unlist(lapply(raster::extract(rl.nlcd.wood,pts.bgris,buffer=2000),mean))
df.bfrat$wood.perc <- unlist(lapply(raster::extract(rl.nlcd.wood,pts.bfrat,buffer=2000),mean))
df.BPEN$wood.perc <- unlist(lapply(raster::extract(rl.nlcd.wood,pts.BPEN,buffer=2000),mean))
df.BIMP$wood.perc <- unlist(lapply(raster::extract(rl.nlcd.wood,pts.BIMP,buffer=2000),mean))
df.BGRIS$wood.perc <- unlist(lapply(raster::extract(rl.nlcd.wood,pts.BGRIS,buffer=2000),mean))
df.BFRAT$wood.perc <- unlist(lapply(raster::extract(rl.nlcd.wood,pts.BFRAT,buffer=2000),mean))
df.quad$wood.perc <- unlist(lapply(raster::extract(rl.nlcd.wood,quad.bimp,buffer=2000),mean)) #2km buffer around survey location, use maximum likelihood to estimate buffer parameter if you are concerned

# Ag percentage
df.bpen$ag.perc <- unlist(lapply(raster::extract(rl.nlcd.ag,pts.bpen,buffer=2000),mean))
df.bimp$ag.perc <- unlist(lapply(raster::extract(rl.nlcd.ag,pts.bimp,buffer=2000),mean))
df.bgris$ag.perc <- unlist(lapply(raster::extract(rl.nlcd.ag,pts.bgris,buffer=2000),mean))
df.bfrat$ag.perc <- unlist(lapply(raster::extract(rl.nlcd.ag,pts.bfrat,buffer=2000),mean))
df.BPEN$ag.perc <- unlist(lapply(raster::extract(rl.nlcd.ag,pts.BPEN,buffer=2000),mean))
df.BIMP$ag.perc <- unlist(lapply(raster::extract(rl.nlcd.ag,pts.BIMP,buffer=2000),mean))
df.BGRIS$ag.perc <- unlist(lapply(raster::extract(rl.nlcd.ag,pts.BGRIS,buffer=2000),mean))
df.BFRAT$ag.perc <- unlist(lapply(raster::extract(rl.nlcd.ag,pts.BFRAT,buffer=2000),mean))
df.quad$ag.perc <- unlist(lapply(raster::extract(rl.nlcd.ag,quad.bimp,buffer=2000),mean)) #2km buffer around survey location, use maximum likelihood to estimate buffer parameter if you are concerned

# Precip
df.bpen$precip <- unlist(lapply(raster::extract(precip,pts.bpen,buffer=2000),mean))
df.bimp$precip <- unlist(lapply(raster::extract(precip,pts.bimp,buffer=2000),mean))
df.bgris$precip <- unlist(lapply(raster::extract(precip,pts.bgris,buffer=2000),mean))
df.bfrat$precip <- unlist(lapply(raster::extract(precip,pts.bfrat,buffer=2000),mean))
df.BPEN$precip <- unlist(lapply(raster::extract(precip,pts.BPEN,buffer=2000),mean))
df.BIMP$precip <- unlist(lapply(raster::extract(precip,pts.BIMP,buffer=2000),mean))
df.BGRIS$precip <- unlist(lapply(raster::extract(precip,pts.BGRIS,buffer=2000),mean))
df.BFRAT$precip <- unlist(lapply(raster::extract(precip,pts.BFRAT,buffer=2000),mean))
df.quad$precip <- unlist(lapply(raster::extract(precip,quad.bimp,buffer=2000),mean)) #2km buffer around survey location, use maximum likelihood to estimate buffer parameter if you are concerned

# Maximum temp
df.bpen$tmax <- unlist(lapply(raster::extract(tmax,pts.bpen,buffer=2000),mean))
df.bimp$tmax <- unlist(lapply(raster::extract(tmax,pts.bimp,buffer=2000),mean))
df.bgris$tmax <- unlist(lapply(raster::extract(tmax,pts.bgris,buffer=2000),mean))
df.bfrat$tmax <- unlist(lapply(raster::extract(tmax,pts.bfrat,buffer=2000),mean))
df.BPEN$tmax <- unlist(lapply(raster::extract(tmax,pts.BPEN,buffer=2000),mean))
df.BIMP$tmax <- unlist(lapply(raster::extract(tmax,pts.BIMP,buffer=2000),mean))
df.BGRIS$tmax <- unlist(lapply(raster::extract(tmax,pts.BGRIS,buffer=2000),mean))
df.BFRAT$tmax <- unlist(lapply(raster::extract(tmax,pts.BFRAT,buffer=2000),mean))
df.quad$tmax <- unlist(lapply(raster::extract(tmax,quad.bimp,buffer=2000),mean)) #2km buffer around survey location, use maximum likelihood to estimate buffer parameter if you are concerned

# Minimum temp
df.bpen$tmin <- unlist(lapply(raster::extract(tmin,pts.bpen,buffer=2000),mean))
df.bimp$tmin <- unlist(lapply(raster::extract(tmin,pts.bimp,buffer=2000),mean))
df.bgris$tmin <- unlist(lapply(raster::extract(tmin,pts.bgris,buffer=2000),mean))
df.bfrat$tmin <- unlist(lapply(raster::extract(tmin,pts.bfrat,buffer=2000),mean))
df.BPEN$tmin <- unlist(lapply(raster::extract(tmin,pts.BPEN,buffer=2000),mean))
df.BIMP$tmin <- unlist(lapply(raster::extract(tmin,pts.BIMP,buffer=2000),mean))
df.BGRIS$tmin <- unlist(lapply(raster::extract(tmin,pts.BGRIS,buffer=2000),mean))
df.BFRAT$tmin <- unlist(lapply(raster::extract(tmin,pts.BFRAT,buffer=2000),mean))
df.quad$tmin <- unlist(lapply(raster::extract(tmin,quad.bimp,buffer=2000),mean)) #2km buffer around survey location, use maximum likelihood to estimate buffer parameter if you are concerned

# Annual mean temp
df.bpen$tmean <- unlist(lapply(raster::extract(tmean,pts.bpen,buffer=2000),mean))
df.bimp$tmean <- unlist(lapply(raster::extract(tmean,pts.bimp,buffer=2000),mean))
df.bgris$tmean <- unlist(lapply(raster::extract(tmean,pts.bgris,buffer=2000),mean))
df.bfrat$tmean <- unlist(lapply(raster::extract(tmean,pts.bfrat,buffer=2000),mean))
df.BPEN$tmean <- unlist(lapply(raster::extract(tmean,pts.BPEN,buffer=2000),mean))
df.BIMP$tmean <- unlist(lapply(raster::extract(tmean,pts.BIMP,buffer=2000),mean))
df.BGRIS$tmean <- unlist(lapply(raster::extract(tmean,pts.BGRIS,buffer=2000),mean))
df.BFRAT$tmean <- unlist(lapply(raster::extract(tmean,pts.BFRAT,buffer=2000),mean))
df.quad$tmean <- unlist(lapply(raster::extract(tmean,quad.bimp,buffer=2000),mean)) #2km buffer around survey location, use maximum likelihood to estimate buffer parameter if you are concerned

end_time <- Sys.time()
elapsed_time <- end_time - start_time
print(elapsed_time)

```



# Make presence column, Join into final dataset

```{r}
# presence column
df.bpen$pa <- 1
df.bimp$pa <- 1
df.bgris$pa <- 1
df.bfrat$pa <- 1
df.BPEN$pa <- 1
df.BIMP$pa <- 1
df.BGRIS$pa <- 1
df.BFRAT$pa <- 1
df.quad$pa <- rep(0, length(df.quad$grass.perc))

colnames(df.quad) <- c("X", "Y", "grass.perc", "dev.perc", "wet.perc",
                       "wood.perc", "ag.perc", "precip", "tmax", "tmin",
                       "tmean", "pa")

colnames(df.bpen) <- c("X", "Y", "grass.perc", "dev.perc", "wet.perc",
                       "wood.perc", "ag.perc", "precip", "tmax", "tmin",
                       "tmean", "pa")

colnames(df.bimp) <- c("X", "Y", "grass.perc", "dev.perc", "wet.perc",
                       "wood.perc", "ag.perc", "precip", "tmax", "tmin",
                       "tmean", "pa")

colnames(df.bgris) <- c("X", "Y", "grass.perc", "dev.perc", "wet.perc",
                       "wood.perc", "ag.perc", "precip", "tmax", "tmin",
                       "tmean", "pa")

colnames(df.bfrat) <- c("X", "Y", "grass.perc", "dev.perc", "wet.perc",
                       "wood.perc", "ag.perc", "precip", "tmax", "tmin",
                       "tmean", "pa")

colnames(df.BPEN) <- c("X", "Y", "grass.perc", "dev.perc", "wet.perc",
                       "wood.perc", "ag.perc", "precip", "tmax", "tmin",
                       "tmean", "pa")

colnames(df.BIMP) <- c("X", "Y", "grass.perc", "dev.perc", "wet.perc",
                       "wood.perc", "ag.perc", "precip", "tmax", "tmin",
                       "tmean", "pa")

colnames(df.BGRIS) <- c("X", "Y", "grass.perc", "dev.perc", "wet.perc",
                       "wood.perc", "ag.perc", "precip", "tmax", "tmin",
                       "tmean", "pa")

colnames(df.BFRAT) <- c("X", "Y", "grass.perc", "dev.perc", "wet.perc",
                       "wood.perc", "ag.perc", "precip", "tmax", "tmin",
                       "tmean", "pa")



# Join into final datasets
df.bpen.ipp <- rbind(df.bpen, df.quad)
df.bimp.ipp <- rbind(df.bimp, df.quad)
df.bgris.ipp <- rbind(df.bgris, df.quad)
df.bfrat.ipp <- rbind(df.bfrat, df.quad)
df.BPEN.ipp <- rbind(df.BPEN, df.quad)
df.BIMP.ipp <- rbind(df.BIMP, df.quad)
df.BGRIS.ipp <- rbind(df.BGRIS, df.quad)
df.BFRAT.ipp <- rbind(df.BFRAT, df.quad)

# Add weights
wt <- 1e-4
df.bpen.ipp$wt <- wt
df.bimp.ipp$wt <- wt
df.bgris.ipp$wt <- wt
df.bfrat.ipp$wt <- wt
df.BPEN.ipp$wt <- wt
df.BIMP.ipp$wt <- wt
df.BGRIS.ipp$wt <- wt
df.BFRAT.ipp$wt <- wt


# Save df for easier use
write.csv(df.bpen.ipp, "df.bpen.ipp.csv")
write.csv(df.bimp.ipp, "df.bimp.ipp.csv")
write.csv(df.bgris.ipp, "df.bgris.ip1p.csv")
write.csv(df.bfrat.ipp, "df.bfrat.ipp.csv")
write.csv(df.BPEN.ipp, "df.bpen.bba.ipp.csv")
write.csv(df.BIMP.ipp, "df.bimp.bba.ipp.csv")
write.csv(df.BFRAT.ipp, "df.bfrat.bba.ipp.csv")
write.csv(df.BGRIS.ipp, "df.bgris.bba.ipp.csv")


# Read df for easier use
df.bimp.ipp <- read.csv("C:/Users/dyhol/OneDrive - Kansas State University/BBSDM/BBSDM_KS_2022/df.bimp.ipp.csv")
df.bpen.ipp <- read.csv("C:/Users/dyhol/OneDrive - Kansas State University/BBSDM/BBSDM_KS_2022/df.bpen.ipp.csv")
df.bgris.ipp <- read.csv("C:/Users/dyhol/OneDrive - Kansas State University/BBSDM/BBSDM_KS_2022/df.bgris.ip1p.csv")
df.bfrat.ipp <- read.csv("C:/Users/dyhol/OneDrive - Kansas State University/BBSDM/BBSDM_KS_2022/df.bfrat.ipp.csv")
df.BIMP.ipp <- read.csv("C:/Users/dyhol/OneDrive - Kansas State University/BBSDM/BBSDM_KS_2022/df.bimp.bba.ipp.csv")
df.BPEN.ipp <- read.csv("C:/Users/dyhol/OneDrive - Kansas State University/BBSDM/BBSDM_KS_2022/df.bpen.bba.ipp.csv")
df.BGRIS.ipp <- read.csv("C:/Users/dyhol/OneDrive - Kansas State University/BBSDM/BBSDM_KS_2022/df.bgris.bba.ipp.csv")
df.BFRAT.ipp <- read.csv("C:/Users/dyhol/OneDrive - Kansas State University/BBSDM/BBSDM_KS_2022/df.bfrat.bba.ipp.csv")

```

# Making Figure X
```{r warning = FALSE}
#Convert to sf
sf.bimp.ipp <- st_as_sf(df.bimp.ipp, coords = c("X","Y"))
sf.bpen.ipp <- st_as_sf(df.bpen.ipp, coords = c("X", "Y"))
sf.BPEN.ipp <- st_as_sf(df.BPEN.ipp, coords = c("X", "Y"))
sf.BIMP.ipp <- st_as_sf(df.BIMP.ipp, coords = c("X", "Y"))

sf.bgris.ipp <- st_as_sf(df.bgris.ipp, coords = c("X","Y"))
sf.bfrat.ipp <- st_as_sf(df.bfrat.ipp, coords = c("X", "Y"))
sf.BGRIS.ipp <- st_as_sf(df.BGRIS.ipp, coords = c("X", "Y"))
sf.BFRAT.ipp <- st_as_sf(df.BFRAT.ipp, coords = c("X", "Y"))

ggplot() +
  geom_sf(data = sf.bimp.ipp, aes(color = pa)) +
  scale_color_viridis_c(option = "D")

par(mfrow = c(2,2))
plot(sf.kansas, main = "B.impatiens occurrence records and \nrandomly-generated quadratures")
plot(quad, add = T)
plot(sf.bimp.ipp[sf.bimp.ipp$pa == 1,], pch = 3, col = "red", add = T)
plot(sf.kansas, main = "B.pensylvanicus occurrence records and \nrandomly-generated quadratures")
plot(quad, add = T)
plot(sf.bpen.ipp[sf.bpen.ipp$pa == 1,], pch = 3, col = "red", add = T)
plot(sf.kansas, main = "B.impatiens survey records and \nrandomly-generated quadratures")
plot(quad, add = T)
plot(sf.BIMP.ipp[sf.BIMP.ipp$pa == 1,], pch = 3, col = "red", add = T)
plot(sf.kansas, main = "B.pensylvanicus survey records and \nrandomly-generated quadratures")
plot(quad, add = T)
plot(sf.BPEN.ipp[sf.BPEN.ipp$pa == 1,], pch = 3, col = "red", add = T)

```



# Model/variable selection -- B. impatiens from occurrence records
```{r}

# The formula with all covariates
f1 <- pa/wt ~ grass.perc + dev.perc + wet.perc + wood.perc + ag.perc + precip + 
  tmax + tmin + tmean

# GLM
ipp <- glm(f1, data = df.bimp.ipp, family = poisson, weights = wt)

summary(ipp)


# Adding spline on location terms
f2 <- update(f1, ~ . + s(X,Y, bs = "gp", k = 200))

ipp <- gam(f2, data = df.bimp.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp)


# Backward elimination: grass.perc
f3 <- update(f2, ~ . + s(X,Y, bs = "gp", k = 200) - grass.perc) 

ipp <- gam(f3, data = df.bimp.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp)

# Backward elimination: tmax
f4 <- update(f3, ~ . + s(X,Y, bs = "gp", k = 200) - tmean) 

ipp <- gam(f4, data = df.bimp.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp)


# Backward elimination: tmin
f5 <- update(f4, ~ . + s(X,Y, bs = "gp", k = 200) - precip) 

ipp <- gam(f5, data = df.bimp.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp)


# Backward elimination: precip
f6 <- update(f5, ~ . + s(X,Y, bs = "gp", k = 200) - wood.perc) 

ipp <- gam(f6, data = df.bimp.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp)


# Backward elimination: wet.perc (FINAL MODEL)
f7 <- update(f6, ~ . + s(X,Y, bs = "gp", k = 200) - wet.perc) 

ipp.bimp <- gam(f7, data = df.bimp.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.bimp) #sig = dev, wood, ag(-), tmean
```


# Model/variable selection -- B. pensylvanicus from occurence records
```{r}
# The formula with all covariates
f1.bpen <- pa/wt ~ grass.perc + dev.perc + wet.perc + wood.perc + ag.perc + precip + 
  tmax + tmin + tmean

# GLM
ipp.bpen <- glm(f1.bpen, data = df.bpen.ipp, family = poisson, weights = wt)

summary(ipp.bpen)


# Adding spline on location terms
f2.bpen <- update(f1.bpen, ~ . + s(X,Y, bs = "gp", k = 200))

ipp.bpen <- gam(f2.bpen, data = df.bpen.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.bpen)


# Backward elimination: 
f3.bpen <- update(f2.bpen, ~ . + s(X,Y, bs = "gp", k = 200) - tmean) 

ipp.bpen <- gam(f3.bpen, data = df.bpen.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.bpen)

# Backward elimination: 
f4.bpen <- update(f3.bpen, ~ . + s(X,Y, bs = "gp", k = 200) - ag.perc) 

ipp.bpen <- gam(f4.bpen, data = df.bpen.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.bpen)

# Backward elimination:
f5.bpen <- update(f4.bpen, ~ . + s(X,Y, bs = "gp", k = 200) - precip) 

ipp.bpen <- gam(f5.bpen, data = df.bpen.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.bpen)

# Backward elimination:
f6.bpen <- update(f5.bpen, ~ . + s(X,Y, bs = "gp", k = 200) - tmax) 

ipp.bpen <- gam(f6.bpen, data = df.bpen.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.bpen)

# Backward elimination:
f7.bpen <- update(f6.bpen, ~ . + s(X,Y, bs = "gp", k = 200) - grass.perc) 

ipp.bpen <- gam(f7.bpen, data = df.bpen.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.bpen) #sig = dev, wet, wood, tmean


```


# Model/variable selection -- B. impatiens from survey records
```{r}
# The formula with all covariates
f1.BIMP <- pa/wt ~ grass.perc + dev.perc + wet.perc + wood.perc + ag.perc + precip + 
  tmax + tmin + tmean

# GLM
ipp.BIMP <- glm(f1.BIMP, data = df.BIMP.ipp, family = poisson, weights = wt)

summary(ipp.BIMP)


# Adding spline on location terms
f2.BIMP <- update(f1.BIMP, ~ . + s(X,Y, bs = "gp", k = 100))

ipp.BIMP <- gam(f2.BIMP, data = df.BIMP.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.BIMP)


# Backward elimination: 
f3.BIMP <- update(f2.BIMP, ~ . + s(X,Y, bs = "gp", k = 100) - grass.perc) 

ipp.BIMP <- gam(f3.BIMP, data = df.BIMP.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.BIMP)

# Backward elimination: 
f4.BIMP <- update(f3.BIMP, ~ . + s(X,Y, bs = "gp", k = 100) - ag.perc) 

ipp.BIMP <- gam(f4.BIMP, data = df.BIMP.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.BIMP)

# Backward elimination: 
f5.BIMP <- update(f4.BIMP, ~ . + s(X,Y, bs = "gp", k = 100) - wet.perc) 

ipp.BIMP <- gam(f5.BIMP, data = df.BIMP.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.BIMP) #sig = dev, wood, tmax
 

# Backward elimination: 
f6.BIMP <- update(f5.BIMP, ~ . + s(X,Y, bs = "gp", k = 100) - precip) 

ipp.BIMP <- gam(f6.BIMP, data = df.BIMP.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.BIMP)


# Backward elimination: 
f7.BIMP <- update(f6.BIMP, ~ . + s(X,Y, bs = "gp", k = 100) - tmean) 

ipp.BIMP <- gam(f7.BIMP, data = df.BIMP.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.BIMP)


# Backward elimination: 
f8.BIMP <- update(f7.BIMP, ~ . + s(X,Y, bs = "gp", k = 100) - tmin) 

ipp.BIMP <- gam(f8.BIMP, data = df.BIMP.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.BIMP)


# Backward elimination: 
f9.BIMP <- update(f8.BIMP, ~ . + s(X,Y, bs = "gp", k = 100) - dev.perc) 

ipp.BIMP <- gam(f9.BIMP, data = df.BIMP.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.BIMP)


# Backward elimination: 
f10.BIMP <- update(f9.BIMP, ~ . + s(X,Y, bs = "gp", k = 100) - wood.perc) 

ipp.BIMP <- gam(f10.BIMP, data = df.BIMP.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.BIMP)

```


# Model/ variable selection -- B. pensylvanicus from survey records
```{r}
# The formula with all covariates
f1.BPEN <- pa/wt ~ grass.perc + dev.perc + wet.perc + wood.perc + ag.perc + precip + 
  tmax + tmin + tmean

# GLM
ipp.BPEN <- glm(f1.BPEN, data = df.BPEN.ipp, family = poisson, weights = wt)

summary(ipp.BPEN)


# Adding spline on location terms
f2.BPEN <- update(f1.BPEN, ~ . + s(X,Y, bs = "gp"))

ipp.BPEN <- gam(f2.BPEN, data = df.BPEN.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.BPEN)


# Backward elimination: 
f3.BPEN <- update(f2.BPEN, ~ . + s(X,Y, bs = "gp") - grass.perc) 

ipp.BPEN <- gam(f3.BPEN, data = df.BPEN.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.BPEN)

# Backward elimination: 
f4.BPEN <- update(f3.BPEN, ~ . + s(X,Y, bs = "gp") - dev.perc) 

ipp.BPEN <- gam(f4.BPEN, data = df.BPEN.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.BPEN)

# Backward elimination: 
f5.BPEN <- update(f4.BPEN, ~ . + s(X,Y, bs = "gp") - wet.perc) 

ipp.BPEN <- gam(f5.BPEN, data = df.BPEN.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.BPEN)

# Backward elimination: 
f6.BPEN <- update(f5.BPEN, ~ . + s(X,Y, bs = "gp", k = 200) - grass.perc) 

ipp.BPEN <- gam(f6.BPEN, data = df.BPEN.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.BPEN)

# Backward elimination: 
f7.BPEN <- update(f6.BPEN, ~ . + s(X,Y, bs = "gp", k = 200) - tmin) 

ipp.BPEN <- gam(f7.BPEN, data = df.BPEN.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.BPEN) #sig = ag(-), tmean


```


# Model/ variable selection -- B. griseocollis from occurrence records
```{r}
# The formula with all covariates
f1.bgris <- pa/wt ~ grass.perc + dev.perc + wet.perc + wood.perc + ag.perc + precip + 
  tmax + tmin + tmean

# GLM
ipp.bgris <- glm(f1.bgris, data = df.bgris.ipp, family = poisson, weights = wt)

summary(ipp.bgris)


# Adding spline on location terms
f2.bgris <- update(f1.bgris, ~ . + s(X,Y, bs = "gp", k = 200))

ipp.bgris <- gam(f2.bgris, data = df.bgris.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.bgris)


# Backward elimination: 
f3.bgris <- update(f2.bgris, ~ . + s(X,Y, bs = "gp", k = 200) - dev.perc) 

ipp.bgris <- gam(f3.bgris, data = df.bgris.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.bgris) 


# Backward elimination: 
f4.bgris <- update(f3.bgris, ~ . + s(X,Y, bs = "gp", k = 200) - wet.perc) 

ipp.bgris <- gam(f4.bgris, data = df.bgris.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.bgris) 


# Backward elimination: 
f5.bgris <- update(f4.bgris, ~ . + s(X,Y, bs = "gp", k = 200) - wood.perc) 

ipp.bgris <- gam(f5.bgris, data = df.bgris.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.bgris) 


# Backward elimination: 
f6.bgris <- update(f5.bgris, ~ . + s(X,Y, bs = "gp", k = 200) - tmax) 

ipp.bgris <- gam(f6.bgris, data = df.bgris.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.bgris) 


# Backward elimination: 
f7.bgris <- update(f6.bgris, ~ . + s(X,Y, bs = "gp", k = 200) - tmin) 

ipp.bgris <- gam(f7.bgris, data = df.bgris.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.bgris)  #sig = grass(-), ag(-), precip (+), tmean (+)


```

# Model/ variable selection -- B. fraternus from occurrence records
```{r}
# The formula with all covariates
f1.bfrat <- pa/wt ~ grass.perc + dev.perc + wet.perc + wood.perc + ag.perc + precip + 
  tmax + tmin + tmean

# GLM
ipp.bfrat <- glm(f1.bfrat, data = df.bfrat.ipp, family = poisson, weights = wt)

summary(ipp.bfrat)


# Adding spline on location terms
f2.bfrat <- update(f1.bfrat, ~ . + s(X,Y, bs = "gp", k = 200))

ipp.bfrat <- gam(f2.bfrat, data = df.bfrat.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.bfrat)


# Backward elimination: 
f3.bfrat <- update(f2.bfrat, ~ . + s(X,Y, bs = "gp", k = 200) - wet.perc)

ipp.bfrat <- gam(f3.bfrat, data = df.bfrat.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.bfrat)

# Backward elimination: 
f4.bfrat <- update(f3.bfrat, ~ . + s(X,Y, bs = "gp", k = 200) - wood.perc) 

ipp.bfrat <- gam(f4.bfrat, data = df.bfrat.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.bfrat)

# Backward elimination: 
f5.bfrat <- update(f4.bfrat, ~ . + s(X,Y, bs = "gp", k = 200) - dev.perc) 

ipp.bfrat <- gam(f5.bfrat, data = df.bfrat.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.bfrat)

# Backward elimination: 
f6.bfrat <- update(f5.bfrat, ~ . + s(X,Y, bs = "gp", k = 200) - tmin) 

ipp.bfrat <- gam(f6.bfrat, data = df.bfrat.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.bfrat)

# Backward elimination: 
f7.bfrat <- update(f6.bfrat, ~ . + s(X,Y, bs = "gp", k = 200) - tmax) 

ipp.bfrat <- gam(f7.bfrat, data = df.bfrat.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.bfrat) 


# Backward elimination: 
f8.bfrat <- update(f7.bfrat, ~ . + s(X,Y, bs = "gp", k = 200) - tmean) 

ipp.bfrat <- gam(f8.bfrat, data = df.bfrat.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.bfrat) 

```

# Model/ variable selection -- B. griseocollis from survey records
```{r}
# The formula with all covariates
f1.BGRIS <- pa/wt ~ grass.perc + dev.perc + wet.perc + wood.perc + ag.perc + precip + 
  tmax + tmin + tmean

# GLM
ipp.BGRIS <- glm(f1.BGRIS, data = df.BGRIS.ipp, family = poisson, weights = wt)

summary(ipp.BGRIS)


# Adding spline on location terms
f2.BGRIS <- update(f1.BGRIS, ~ . + s(X,Y, bs = "gp", k = 200))

ipp.BGRIS <- gam(f2.BGRIS, data = df.BGRIS.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.BGRIS)


# Backward elimination: 
f3.BGRIS <- update(f2.BGRIS, ~ . + s(X,Y, bs = "gp", k = 200) - wet.perc) 

ipp.BGRIS <- gam(f3.BGRIS, data = df.BGRIS.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.BGRIS)
AIC(ipp.BGRIS)

# Backward elimination: 
f4.BGRIS <- update(f3.BGRIS, ~ . + s(X,Y, bs = "gp", k = 200) - precip)

ipp.BGRIS <- gam(f4.BGRIS, data = df.BGRIS.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.BGRIS)
AIC(ipp.BGRIS) # 307.0223

# Backward elimination: 
f5.BGRIS <- update(f4.BGRIS, ~ . + s(X,Y, bs = "gp", k = 200) - tmin) 

ipp.BGRIS <- gam(f5.BGRIS, data = df.BGRIS.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.BGRIS)
AIC(ipp.BGRIS) # 306.5185

# Backward elimination: 
f6.BGRIS <- update(f5.BGRIS, ~ . + s(X,Y, bs = "gp", k = 200) - tmax) 

ipp.BGRIS <- gam(f6.BGRIS, data = df.BGRIS.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.BGRIS)
AIC(ipp.BGRIS) # 307.2605

# Backward elimination: 
f7.BGRIS <- update(f6.BGRIS, ~ . + s(X,Y, bs = "gp", k = 200) - tmean) 

ipp.BGRIS <- gam(f7.BGRIS, data = df.BGRIS.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.BGRIS)
AIC(ipp.BGRIS) # 301.2894

# Backward elimination: 
f8.BGRIS <- update(f7.BGRIS, ~ . + s(X,Y, bs = "gp", k = 200) - dev.perc) 

ipp.BGRIS <- gam(f8.BGRIS, data = df.BGRIS.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.BGRIS)
#AIC(ipp.BGRIS) 304.3393
#sig = wet, ag(-), precip

```

# Model/ variable selection -- B. fraternus from survey records
```{r}
# The formula with all covariates
f1.BFRAT <- pa/wt ~ grass.perc + dev.perc + wet.perc + wood.perc + ag.perc + precip + 
  tmax + tmin + tmean

# GLM
ipp.BFRAT <- glm(f1.BFRAT, data = df.BFRAT.ipp, family = poisson, weights = wt)

summary(ipp.BFRAT)


# Adding spline on location terms
f2.BFRAT <- update(f1.BFRAT, ~ . + s(X,Y, bs = "gp", k = 100))

ipp.BFRAT <- gam(f2.BFRAT, data = df.BFRAT.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.BFRAT)


# Backward elimination: 
f3.BFRAT <- update(f2.BFRAT, ~ . + s(X,Y, bs = "gp", k = 100) - dev.perc) 

ipp.BFRAT <- gam(f3.BFRAT, data = df.BFRAT.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.BFRAT) 

# Backward elimination: 
f4.BFRAT <- update(f3.BFRAT, ~ . + s(X,Y, bs = "gp", k = 100) - wet.perc) 

ipp.BFRAT <- gam(f4.BFRAT, data = df.BFRAT.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.BFRAT) 

# Backward elimination: 
f5.BFRAT <- update(f4.BFRAT, ~ . + s(X,Y, bs = "gp", k = 100) - precip) 

ipp.BFRAT <- gam(f5.BFRAT, data = df.BFRAT.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.BFRAT)

# Backward elimination: 
f6.BFRAT <- update(f5.BFRAT, ~ . + s(X,Y, bs = "gp", k = 100) - tmax) 

ipp.BFRAT <- gam(f6.BFRAT, data = df.BFRAT.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.BFRAT) 

# Backward elimination: 
f7.BFRAT <- update(f6.BFRAT, ~ . + s(X,Y, bs = "gp", k = 100) - tmin) 

ipp.BFRAT <- gam(f7.BFRAT, data = df.BFRAT.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.BFRAT) 

# Backward elimination: 
f8.BFRAT <- update(f7.BFRAT, ~ . + s(X,Y, bs = "gp", k = 100) - grass.perc) 

ipp.BFRAT <- gam(f8.BFRAT, data = df.BFRAT.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.BFRAT) 


# Backward elimination
f9.BFRAT <- update(f8.BFRAT, ~ . + s(X,Y, bs = "gp", k = 100) - tmean) 

ipp.BFRAT <- gam(f9.BFRAT, data = df.BFRAT.ipp, family = poisson,
           weights = wt, method = "REML")

summary(ipp.BFRAT) 

```




# Creating new data for model-fitting
```{r}
#Create empty raster layer, add 
start_time <- Sys.time()
rl.E.y.3k <- raster(,nrow=150,ncols=150,ext=extent(sf.kansas),crs=crs(sf.kansas))
df.pred.3k <- data.frame(long = xyFromCell(rl.E.y.3k,cell=1:length(rl.E.y.3k[]))[,1],
                      lat = xyFromCell(rl.E.y.3k,cell=1:length(rl.E.y.3k[]))[,2])

#Extract same covariates as before from our "new data" quadrats
pred.grass <- unlist(lapply(raster::extract(rl.nlcd.grass,df.pred.3k,buffer=2000),mean))*100 #2km buffer around survey location, use maximum likelihood to estimate buffer parameter if you are concerned

# Dev percentage
pred.dev <- unlist(lapply(raster::extract(rl.nlcd.dev,df.pred.3k,buffer=2000),mean))*100 #2km buffer around survey location, use maximum likelihood to estimate buffer parameter if you are concerned

# Wet percentage
pred.wet <- unlist(lapply(raster::extract(rl.nlcd.wet,df.pred.3k,buffer=2000),mean))*100 #2km buffer around survey location, use maximum likelihood to estimate buffer parameter if you are concerned

# Wood percentage
pred.wood <- unlist(lapply(raster::extract(rl.nlcd.wood,df.pred.3k,buffer=2000),mean))*100 #2km buffer around survey location, use maximum likelihood to estimate buffer parameter if you are concerned

# Ag percentage
pred.ag <- unlist(lapply(raster::extract(rl.nlcd.ag,df.pred.3k,buffer=2000),mean))*100 #2km buffer around survey location, use maximum likelihood to estimate buffer parameter if you are concerned

# Precip
pred.precip <- unlist(lapply(raster::extract(precip,df.pred.3k,buffer=2000),mean)) #2km buffer around survey location, use maximum likelihood to estimate buffer parameter if you are concerned

# Maximum temp
pred.tmax <- unlist(lapply(raster::extract(tmax,df.pred.3k,buffer=2000),mean)) #2km buffer around survey location, use maximum likelihood to estimate buffer parameter if you are concerned

# Minimum temp
pred.tmin <- unlist(lapply(raster::extract(tmin,df.pred.3k,buffer=2000),mean)) #2km buffer around survey location, use maximum likelihood to estimate buffer parameter if you are concerned

# Annual mean temp
pred.tmean <- unlist(lapply(raster::extract(tmean,df.pred.3k,buffer=2000),mean)) #2km buffer around survey location, use maximum likelihood to estimate buffer parameter if you are concerned


df.pred.all3k <- df.pred.3k
df.pred.all3k$dev.perc <- pred.dev
df.pred.all3k$wet.perc <- pred.wet
df.pred.all3k$wood.perc <- pred.wood
df.pred.all3k$ag.perc <- pred.ag
df.pred.all3k$grass.perc <- pred.grass
df.pred.all3k$precip <- pred.precip
df.pred.all3k$tmean <- pred.tmean
df.pred.all3k$tmin <- pred.tmin
df.pred.all3k$tmax <- pred.tmax

end_time <- Sys.time()
elapsed_time <- end_time - start_time
print(elapsed_time)

write.csv(df.pred.all3k, "C:/Users/dyhol/OneDrive - Kansas State University/BBSDM/BBSDM_KS_2022/df.pred.all3k")

df.pred.all3k <- read.csv("C:/Users/dyhol/OneDrive - Kansas State University/BBSDM/BBSDM_KS_2022/df.pred.all3k")

```


# Fit each model with the new data & plot intensity -- B. impatiens from occurrence records

```{r}
# Select only significant covariates for the species & add to df.pred for the species
#            Estimate Std. Error z value Pr(>|z|)    
#(Intercept) -30.8311    15.3684  -2.006 0.044842 *  
#dev.perc      2.1559     0.3730   5.780 7.49e-09 ***
#ag.perc      -2.9202     0.7944  -3.676 0.000237 ***
#tmax          1.1049     0.5444   2.030 0.042381 *  
#tmin          0.9242     0.5023   1.840 0.065753 .  
df.pred.bimp.3k <- df.pred.3k

df.pred.bimp.3k$dev.perc <- df.pred.all3k$dev.perc
df.pred.bimp.3k$ag.perc <- df.pred.all3k$ag.perc
df.pred.bimp.3k$tmax <- df.pred.all3k$tmax
df.pred.bimp.3k$tmin <- df.pred.all3k$tmin

df.pred.bimp.3k$dev.perc <- df.pred.bimp.3k$dev.perc*.01 
df.pred.bimp.3k$ag.perc <- df.pred.bimp.3k$ag.perc*.01 

colnames(df.pred.bimp.3k) <- c("X","Y","dev.perc", "ag.perc", "tmax", "tmin")


rl.E.y.bimp.3k <- rl.E.y.3k
rl.E.y.bimp.3k[] <- exp(c(predict(ipp.bimp,newdata=df.pred.bimp.3k,type="link")))

#image(rl.E.y,axes=FALSE,box=FALSE)
plot(rl.E.y.bimp.3k,axes=FALSE,box=FALSE,xlim=c(-102.3,-94.6),ylim=c(37,40))
plot(sf.kansas,add=TRUE)
plot(pts.bimp, add = TRUE, cex = 0.2)


```



# Fit each model with the new data & plot intensity -- B. pensylvanicus from occurrence records

Significant covariates = dev, wet, wood, tmean

```{r}
# Select only significant covariates for the species & add to df.pred for the species

#            Estimate Std. Error z value Pr(>|z|)    
#(Intercept) -0.05935    1.81287  -0.033   0.9739    
#dev.perc     4.36030    0.35742  12.199  < 2e-16 ***
#wet.perc     3.82763    0.96341   3.973 7.10e-05 ***
#wood.perc    5.42106    0.93700   5.786 7.23e-09 ***
#tmin         0.62644    0.37542   1.669   0.0952 .  


df.pred.bpen.3k <- df.pred.3k

colnames(df.pred.bpen.3k) <- c("X","Y")

df.pred.bpen.3k$dev.perc <- df.pred.all3k$dev.perc
df.pred.bpen.3k$wet.perc <- df.pred.all3k$wet.perc
df.pred.bpen.3k$wood.perc <- df.pred.all3k$wood.perc
df.pred.bpen.3k$tmin <- df.pred.all3k$tmin

df.pred.bpen.3k$dev.perc <- df.pred.bpen.3k$dev.perc*.01 
df.pred.bpen.3k$wood.perc <- df.pred.bpen.3k$wood.perc*.01 
df.pred.bpen.3k$wet.perc <- df.pred.bpen.3k$wet.perc*.01 



rl.E.y.bpen.3k <- rl.E.y.3k
rl.E.y.bpen.3k[] <- exp(c(predict(ipp.bpen,newdata=df.pred.bpen.3k,type="link")))

#image(rl.E.y,axes=FALSE,box=FALSE)
plot(rl.E.y.bpen.3k,axes=FALSE,box=FALSE,xlim=c(-102.3,-94.6),ylim=c(37,40))
plot(sf.kansas,add=TRUE)
plot(pts.bpen, add = TRUE, cex = 0.2)

```


# Fit each model with the new data & plot intensity -- B. impatiens from survey records

Significant covariates = dev, wood, tmax

```{r}
# Select only significant covariates for the species & add to df.pred for the species

#            Estimate Std. Error z value Pr(>|z|)    
#(Intercept) -144.034     65.050  -2.214 0.026816 *  
#tmax           5.339      1.571   3.399 0.000677 ***


df.pred.BIMP.3k <- df.pred.3k
colnames(df.pred.BIMP.3k) <- c("X","Y")

df.pred.BIMP.3k$grass.perc <- df.pred.all3k$grass.perc
df.pred.BIMP.3k$dev.perc <- df.pred.all3k$dev.perc
df.pred.BIMP.3k$wet.perc <- df.pred.all3k$wet.perc
df.pred.BIMP.3k$wood.perc <- df.pred.all3k$wood.perc
df.pred.BIMP.3k$ag.perc <- df.pred.all3k$ag.perc

df.pred.BIMP.3k$tmax <- df.pred.all3k$tmax
df.pred.BIMP.3k$tmin <- df.pred.all3k$tmin
df.pred.BIMP.3k$tmean <- df.pred.all3k$tmean
df.pred.BIMP.3k$precip <- df.pred.all3k$precip

df.pred.BIMP.3k$grass.perc <- df.pred.BIMP.3k$grass.perc*.01 
df.pred.BIMP.3k$wood.perc <- df.pred.BIMP.3k$wood.perc*.01 
df.pred.BIMP.3k$dev.perc <- df.pred.BIMP.3k$dev.perc*.01 
df.pred.BIMP.3k$wet.perc <- df.pred.BIMP.3k$wet.perc*.01 
df.pred.BIMP.3k$ag.perc <- df.pred.BIMP.3k$ag.perc*.01 



rl.E.y.BIMP.3k <- rl.E.y.3k
rl.E.y.BIMP.3k[] <- exp(c(predict(ipp.BIMP,newdata=df.pred.BIMP.3k,type="link")))

#image(rl.E.y,axes=FALSE,box=FALSE)
plot(rl.E.y.BIMP.3k,axes=FALSE,box=FALSE,xlim=c(-102.3,-94.6),ylim=c(37,40))
plot(sf.kansas,add=TRUE)
plot(pts.BIMP, add = TRUE)

```




# Fit each model with the new data & plot intensity -- B. pensylvanicus from survey records

Significant covariates = grass, wet, wood, 

```{r}
# Select only significant covariates for the species & add to df.pred for the species

#            Estimate Std. Error z value Pr(>|z|)    
#(Intercept)  -28.149     28.366  -0.992 0.321022    
#wood.perc      7.318      1.950   3.752 0.000175 ***
#ag.perc       -5.216      0.967  -5.394  6.9e-08 ***
#tmax          -2.544      1.385  -1.838 0.066104 .  
#tmean          3.601      1.840   1.957 0.050374 .  


df.pred.BPEN.3k <- df.pred.3k
colnames(df.pred.BPEN.3k) <- c("X","Y")

df.pred.BPEN.3k$wood.perc <- df.pred.all3k$wood.perc
df.pred.BPEN.3k$ag.perc <- df.pred.all3k$ag.perc
df.pred.BPEN.3k$tmean <- df.pred.all3k$tmean
df.pred.BPEN.3k$tmax <- df.pred.all3k$tmax



df.pred.BPEN.3k$wood.perc <- df.pred.BPEN.3k$wood.perc*.01
df.pred.BPEN.3k$ag.perc <- df.pred.BPEN.3k$ag.perc*.01 


#colnames(df.pred.BPEN.3k) <- c("X","Y","ag.perc", "tmean")


rl.E.y.BPEN.3k <- rl.E.y.3k
rl.E.y.BPEN.3k[] <- exp(c(predict(ipp.BPEN,newdata=df.pred.BPEN.3k,type="link")))

#image(rl.E.y,axes=FALSE,box=FALSE)
plot(rl.E.y.BPEN.3k,axes=FALSE,box=FALSE,xlim=c(-102.3,-94.6),ylim=c(37,40))
plot(sf.kansas,add=TRUE)
plot(pts.BPEN, add = TRUE, cex = 0.5)


```


# Fit each model with the new data & plot intensity -- B. griseocollis from occurrence records

Significant covariates = grass(-), wet, wood(-), ag(-), precip, tmax(-), tmin(-), tmean

```{r}
# Select only significant covariates for the species & add to df.pred for the species
#              Estimate Std. Error z value Pr(>|z|)    
#(Intercept) -17.826668   8.846803  -2.015   0.0439 *  
#grass.perc   -2.276723   0.432514  -5.264 1.41e-07 ***
#ag.perc      -5.380302   0.515522 -10.437  < 2e-16 ***
#precip        0.003869   0.002350   1.646   0.0997 .  
#tmean         1.454103   0.567639   2.562   0.0104 *  

df.pred.bgris.3k <- df.pred.3k
colnames(df.pred.bgris.3k) <- c("X","Y")

df.pred.bgris.3k$grass.perc <- df.pred.all3k$grass.perc
df.pred.bgris.3k$wet.perc <- df.pred.all3k$wet.perc
df.pred.bgris.3k$wood.perc <- df.pred.all3k$wood.perc
df.pred.bgris.3k$ag.perc <- df.pred.all3k$ag.perc
df.pred.bgris.3k$precip <- df.pred.all3k$precip
df.pred.bgris.3k$tmax <- df.pred.all3k$tmax
df.pred.bgris.3k$tmin <- df.pred.all3k$tmin
df.pred.bgris.3k$tmean <- df.pred.all3k$tmean

df.pred.bgris.3k$grass.perc <- df.pred.bgris.3k$grass.perc*.01 
df.pred.bgris.3k$wet.perc <- df.pred.bgris.3k$wet.perc*.01 
df.pred.bgris.3k$wood.perc <- df.pred.bgris.3k$wood.perc*.01 
df.pred.bgris.3k$ag.perc <- df.pred.bgris.3k$ag.perc*.01 


rl.E.y.bgris.3k <- rl.E.y.3k
rl.E.y.bgris.3k[] <- exp(c(predict(ipp.bgris,newdata=df.pred.bgris.3k,type="link")))

#image(rl.E.y,axes=FALSE,box=FALSE)
plot(rl.E.y.bgris.3k,axes=FALSE,box=FALSE,xlim=c(-102.3,-94.6),ylim=c(37,40))
plot(sf.kansas,add=TRUE)
plot(pts.bgris, add = TRUE, cex = 0.5)


```

# Fit each model with the new data & plot intensity -- B. fraternus from occurrence records

Significant covariates = grass(-), wet, wood(-), ag(-)

```{r}
# Select only significant covariates for the species & add to df.pred for the species

#             Estimate Std. Error z value Pr(>|z|)    
#(Intercept) -0.532838   3.732594  -0.143   0.8865    
#grass.perc  -3.227306   0.657793  -4.906 9.28e-07 ***
#ag.perc     -4.398077   0.736867  -5.969 2.39e-09 ***
#precip       0.005478   0.003121   1.755   0.0792 .  

df.pred.bfrat.3k <- df.pred.3k
colnames(df.pred.bfrat.3k) <- c("X","Y")

df.pred.bfrat.3k$grass.perc <- df.pred.all3k$grass.perc
df.pred.bfrat.3k$ag.perc <- df.pred.all3k$ag.perc
df.pred.bfrat.3k$precip <- df.pred.all3k$precip

df.pred.bfrat.3k$grass.perc <- df.pred.bfrat.3k$grass.perc*.01 
df.pred.bfrat.3k$ag.perc <- df.pred.bfrat.3k$ag.perc*.01 





rl.E.y.bfrat.3k <- rl.E.y.3k
rl.E.y.bfrat.3k[] <- exp(c(predict(ipp.bfrat,newdata=df.pred.bfrat.3k,type="link")))

#image(rl.E.y,axes=FALSE,box=FALSE)
color <- colorRampPalette(rev(brewer.pal(11, "Spectral"))[-1],
                          bias = 1.5)

plot(rl.E.y.bfrat.3k,axes=FALSE,box=FALSE,xlim=c(-102.3,-94.6),ylim=c(37,40))
plot(sf.kansas,add=TRUE)
plot(pts.bfrat, add = TRUE, cex = 0.5)


```


# Fit each model with the new data & plot intensity -- B. griseocollis from survey records

Significant covariates = wet, ag(-), precip

```{r}
# Select only significant covariates for the species & add to df.pred for the species

#            Estimate Std. Error z value Pr(>|z|)    
#(Intercept) -16.5137    18.0535  -0.915  0.36035    
#grass.perc   -1.7387     0.8342  -2.084  0.03713 *  
#wood.perc     6.4848     2.1257   3.051  0.00228 ** 
#ag.perc      -6.6241     1.4135  -4.686 2.78e-06 ***


df.pred.BGRIS.3k <- df.pred.3k
colnames(df.pred.BGRIS.3k) <- c("X","Y")

df.pred.BGRIS.3k$grass.perc <- df.pred.all3k$grass.perc
df.pred.BGRIS.3k$dev.perc <- df.pred.all3k$dev.perc
df.pred.BGRIS.3k$wood.perc <- df.pred.all3k$wood.perc
df.pred.BGRIS.3k$ag.perc <- df.pred.all3k$ag.perc




df.pred.BGRIS.3k$grass.perc <- df.pred.BGRIS.3k$grass.perc*.01 
df.pred.BGRIS.3k$dev.perc <- df.pred.BGRIS.3k$dev.perc*.01 
df.pred.BGRIS.3k$wood.perc <- df.pred.BGRIS.3k$wood.perc*.01 
df.pred.BGRIS.3k$ag.perc <- df.pred.BGRIS.3k$ag.perc*.01 




rl.E.y.BGRIS.3k <- rl.E.y.3k
rl.E.y.BGRIS.3k[] <- exp(c(predict(ipp.BGRIS,newdata=df.pred.BGRIS.3k,type="link")))

#image(rl.E.y,axes=FALSE,box=FALSE)
plot(rl.E.y.BGRIS.3k,axes=FALSE,box=FALSE,xlim=c(-102.3,-94.6),ylim=c(37,40))
plot(sf.kansas,add=TRUE)
plot(pts.BGRIS, add = TRUE, cex = 0.5)


```

# Fit each model with the new data & plot intensity -- B. fraternus from survey records

Significant covariates = wet, ag(-)

```{r}
# Select only significant covariates for the species & add to df.pred for the species

#            Estimate Std. Error z value Pr(>|z|)    
#(Intercept)  -0.9282    27.9055  -0.033  0.97347    
#grass.perc   -1.3754     0.8015  -1.716  0.08615 .  
#wood.perc     9.2053     2.8153   3.270  0.00108 ** 
#ag.perc      -4.4091     1.0175  -4.333 1.47e-05 ***
#tmean        -2.9370     1.7167  -1.711  0.08711 .  


df.pred.BFRAT.3k <- df.pred.3k
colnames(df.pred.BFRAT.3k) <- c("X","Y")


df.pred.BFRAT.3k$grass.perc <- df.pred.all3k$grass.perc*.01
df.pred.BFRAT.3k$wood.perc <- df.pred.all3k$wood.perc*.01
df.pred.BFRAT.3k$precip <- df.pred.all3k$precip
df.pred.BFRAT.3k$ag.perc <- df.pred.all3k$ag.perc*.01
df.pred.BFRAT.3k$tmean <- df.pred.all3k$tmean


rl.E.y.BFRAT.3k <- rl.E.y.3k
rl.E.y.BFRAT.3k[] <- exp(c(predict(ipp.BFRAT,newdata=df.pred.BFRAT.3k,type="link")))

#image(rl.E.y,axes=FALSE,box=FALSE)
plot(rl.E.y.BFRAT.3k,axes=FALSE,box=FALSE,xlim=c(-102.3,-94.6),ylim=c(37,40))
plot(sf.kansas,add=TRUE)
plot(pts.BFRAT, add = TRUE, cex = 0.1)



```

# All intensity function plots side-by-side

```{r}
par(mfrow = c(2,2))
plot(rl.E.y.bimp.3k,axes=FALSE,box=FALSE,xlim=c(-102.3,-94.6),ylim=c(37,40),
     main = "B. impatiens from \noccurence records") 
plot(sf.kansas,add=TRUE)
plot(rl.E.y.bpen.3k,axes=FALSE,box=FALSE,xlim=c(-102.3,-94.6),ylim=c(37,40),
     main = "B. pensylvanicus from \noccurence records")
plot(sf.kansas,add=TRUE)
plot(rl.E.y.BIMP.3k,axes=FALSE,box=FALSE,xlim=c(-102.3,-94.6),ylim=c(37,40),
     main = "B. impatiens from \nsurvey records")
plot(sf.kansas,add=TRUE)
plot(rl.E.y.BPEN.3k,axes=FALSE,box=FALSE,xlim=c(-102.3,-94.6),ylim=c(37,40),
     main = "B. pensylvanicus from \nsurvey records")
plot(sf.kansas,add=TRUE)



par(mfrow = c(2,2))
plot(rl.E.y.bgris.3k,axes=FALSE,box=FALSE,xlim=c(-102.3,-94.6),ylim=c(37,40),
     main = "B. griseocollis from \noccurence records") 
plot(sf.kansas,add=TRUE)
plot(rl.E.y.bfrat.3k,axes=FALSE,box=FALSE,xlim=c(-102.3,-94.6),ylim=c(37,40),
     main = "B. fraternus from \noccurence records")
plot(sf.kansas,add=TRUE)
plot(rl.E.y.BGRIS.3k,axes=FALSE,box=FALSE,xlim=c(-102.3,-94.6),ylim=c(37,40),
     main = "B. griseocollis from \nsurvey records")
plot(sf.kansas,add=TRUE)
plot(rl.E.y.BFRAT.3k,axes=FALSE,box=FALSE,xlim=c(-102.3,-94.6),ylim=c(37,40),
     main = "B. fraternus from \nsurvey records")
plot(sf.kansas,add=TRUE)


```